---
title: "generate biomarker figure"
author: "Amy Anderson"
date: "2024-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load libraries
library(tidyverse)
library(ggplot2)
library(plyr)



# load model results from 'Tsimane cranial lesiona dn adult health.Rmd'
load("./Models/pcls_hemoglobin.rda")
load("./Models/pcls_wbc.rda")
load("./Models/pcls_neutrophils.rda")
load("./Models/pcls_eosinophils.rda")
load("./Models/pcls_lymphocytes.rda")
load("./Models/naturalkillers_mod.rda")
load("./Models/bcells_mod.rda")
load("./Models/total_cd4_mod.rda")
load("./Models/naivecd4_mod.rda")
load("./Models/total_cd8_mod.rda")
load("./Models/naivecd8_mod.rda")
load("./Models/cd4to8_mod.rda")


## Read in data file.
d <- read_csv("./Data/Processed_Data/post_biomarker_models.csv")

```



```{r}
mods_list <- list(hb_history_mod, # list of model outputs
                  wbc_mod,
                  eosin_mod, 
                  neutro_mod, 
                  lympho_mod, 
                  nk_mod, 
                  b_mod,
                  cd4,
                  naivecd4, 
                  cd8,
                  naivecd8, 
                  cd4to8)
# extract population-level effects from each model. 
mod_mains <- map(mods_list, fixef)





# Function to write a table of model results
get.table <- function(x){
  addparentheses <- function(x){paste0("(",x,")")}
  addpercent <- function(x){paste0(x,"%")}
   test <- data.frame()
  for(i in 1:length(x)){
     co_estimate <- round(x[[i]][4,1], 2)
     ph_estimate <- round(x[[i]][5,1], 2)
     colower_ci <- round(x[[i]][4,3], 2)
    coupper_ci <- round(x[[i]][4,4], 2)
    phlower_ci <- round(x[[i]][5,3], 2)
    phupper_ci <- round(x[[i]][5,4], 2)
     coci <- addparentheses(paste(colower_ci, coupper_ci, sep = ", "))
     phci <- addparentheses(paste(phlower_ci, phupper_ci, sep = ", "))
     newrow <- data.frame(t((c(co_estimate, coci, ph_estimate, phci))))
     test <- (rbind.fill(test, newrow))
   }
   return(test)
}




blood.table <- get.table(mod_mains)
rownames(blood.table) <- c("hemoglobin", "WBC", "eosinophils", "neutrophils", "lymphocytes", "natural killer cells", "B cells", "total CD4 cells", "naive CD4 cells", "total CD8 cells", "naive CD8 cells", "CD4/CD8 ratio")

colnames(blood.table) <- c("CO", "95% CI", "PH", "95% CI")

write.csv(blood.table, file = "./Models/hematology_betas.csv")

blood.table
```


```{r}
#### Bespoke Functions for predicting posterior values ####
# Produce posterior predicted values
get.posterior <- function(x){
  post <- as.data.frame(posterior_epred(x, newdata = nd, re_formula = NA))
  names(post) <- c("lesions_absent", "orbital_porosity", "vault_porosity")
  return(post)
}
# add columns for effect size (difference between baseline -- lesions absent -- and orbital/vault lesions present)
add.deltas <- function(x){
  x %>%
    mutate(orbital_delta = orbital_porosity - lesions_absent,
         vault_delta = vault_porosity - lesions_absent)
}
add.delta.percents <- function(x){
  x %>%
    mutate(orbital_delta_percent = (orbital_porosity - lesions_absent) / median(lesions_absent) * 100,
    vault_delta_percent = (vault_porosity - lesions_absent) / median(lesions_absent) * 100) 
}

# new data for posterior predictions (included this as an object inside and not an argument of get.posterior() because all predicted values will use the same data frame -- this one -- for predictions)
age <- rep(median(d$age), 3)
sex <- rep("female", 3)
co <- c(0,1,0)
ph <- c(0,0,1)

nd <- data.frame(age, sex, co, ph)
```



*Calculate distribution of posterior predicted mean values for hematological values, with/without lesions:*  
Predicted mean values and 95% credibility interval for a female at the average sample age, given varying lesion status (no lesions, orbital porosity, cranial vault porosity). Deltas provide an intuitive way to interpret effect size, showing the difference between the distributions of predicted means for individuals with and without lesions. (e.g., the predicted mean B cell count for an average-aged woman in the sample is 30% (3.4-56%) lower if she has orbital roof porosity than if she has no porous cranial lesions). 
```{r}

mods_list <- list(hb_history_mod, # list of model outputs
                  wbc_mod,
                  eosin_mod, 
                  neutro_mod, 
                  lympho_mod, 
                  nk_mod, 
                  b_mod,
                  cd4,
                  naivecd4, 
                  cd8,
                  naivecd8, 
                  cd4to8)

resid_wbc <- sapply(list(d$leuco_log, # calculate median residual WBC count relevant for each model
                  d$leuco_log, 
                  d$wbc_sanseosin, 
                  d$wbc_sansneutro, 
                  d$wbc_sanslympho, 
                  d$wbc_sansnk, 
                  d$wbc_sansb, 
                  d$wbc_sanscd4,
                  d$wbc_sansnaivecd4, 
                  d$wbc_sanscd8,
                  d$wbc_sansnaivecd8, 
                  d$wbc_sanscd48), median, na.rm = T)

# calculate posterior predicted parameters for all hematology models
## general form of function for calculating posterior mean estimates
get.posterior <- function(x, nd){
  post <- as.data.frame(posterior_epred(x, newdata = nd, re_formula = NA))
  names(post) <- c("lesions_absent", "orbital_porosity", "vault_porosity")
  return(post)
}


x = mods_list
i = 1
posterior_loop <- function(x){
    post <- list() # create empty list object to populate
  for(i in seq(x)){
    nd1 <- data.frame(age = rep(median(d$age),3), sex = "female", co = c(0,1,0), ph = c(0,0,1)) # unchanging values for newdata predictions
      nd <- data.frame(cbind(nd1, rep(resid_wbc[i], 3))) # changeable value (residual WBC) of newdata
      resid_names <- c("leuco_log", 
                       "leuco_log", 
                       "wbc_sanseosin", 
                       "wbc_sansneutro", 
                       "wbc_sanslympho", 
                       "wbc_sansnk", 
                       "wbc_sansb", 
                       "wbc_sanscd4",
                       "wbc_sansnaivecd4", 
                       "wbc_sanscd8",
                       "wbc_sansnaivecd8", 
                       "wbc_sanscd48")
  names(nd)[5] <- resid_names[i] # rename the resid_wbc column of nd so it matches the model covariate
  newpost <- get.posterior(x[[i]], nd) # calculate posterior mean estimates for model
  post[[length(post) + 1]] <- newpost # new values from line above become item in list
  }
    assign("post", post, envir = globalenv()) # save the finished list in the global environment
}

# execute the function above:
posterior_loop(mods_list[3]) # output is in the object 'post', a list with 10 data frames
 

# exponentiate all posterior values except for hemoglobin (the only model not run on logged values)
postreal <- post
postreal[2:10] <- lapply(post[c(2:10)], exp)


# calculate difference (delta) between posterior predicted means of outcomes with/without lesions 
blood_post_deltas <- lapply(postreal, FUN = add.delta.percents)
wbcpostdelta <- add.delta.percents(exp(get.posterior(wbc_mod)))
blood_post_deltas[[11]] <- wbcpostdelta
```




Data for forest plot: mean delta and 95% CIs:
```{r}
library(HDInterval)
library(plyr)
library(purrr)
posterior_means <- for(i in seq(blood_post_deltas){
  
},.depth = 1, .f = mean)

# x = blood_post_deltas
# i = 1
delta.table <- function(x){
  deltas <- data.frame()
  variables <- c("hemoglobin", "total leukocytes", "eosinophils", "neutrophils", "lymphocytes", "natural killer cells", "B cells", "total CD4 cells", "naive CD4 cells", "total CD8 cells", "naive CD8 cells", "CD4/CD8 ratio")
  for(i in seq(x)){
  co_delta <- mean(x[[i]][,4])
  co_lower <- as.numeric(paste(round(hdi(blood_post_deltas[[i]][,4])[1], 2)))
  co_upper <- as.numeric(paste(round(hdi(blood_post_deltas[[i]][,4])[2], 2)))
  ph_delta <- mean(x[[i]][,5])
  ph_lower <- as.numeric(paste(round(hdi(blood_post_deltas[[i]][,5])[1], 2)))
  ph_upper <- as.numeric(paste(round(hdi(blood_post_deltas[[i]][,5])[2], 2)))

  newdelta1 <- data.frame(variables[i], co_delta, co_lower, co_upper, "orbit")
  names(newdelta1) <- c("variable", "mean", "lower", "upper", "lesion")
  newdelta2 <- data.frame(variables[i], ph_delta, ph_lower, ph_upper, "vault")
  names(newdelta2) <- c("variable", "mean", "lower", "upper", "lesion")
  deltas <- rbind.fill(deltas, newdelta1, newdelta2)
  }
  deltas
}

forestplot_data <- delta.table(blood_post_deltas)
```


Forest plot of hematological models: % difference in mean predicted values for a median-aged female compared to an age-and-sex-matched individual without cranial lesions:
```{r}
forestplot_data$variable <- factor(forestplot_data$variable, levels=rev(c("hemoglobin", "total leukocytes", "eosinophils", "neutrophils", "lymphocytes", "natural killer cells", "B cells", "total CD4 cells", "naive CD4 cells", "total CD8 cells", "naive CD8 cells", "CD4/CD8 ratio")))

bloodforest <- ggplot(data=forestplot_data, aes(x=variable, y=mean, ymin=lower, ymax=upper)) +
        geom_pointrange(aes(color = lesion), position = position_dodge(width = 0.5)) + 
        scale_color_manual(values = c("tomato", "slateblue")) + 
        geom_hline(yintercept=1, lty=2) +  # add a dotted line at x=1 after flip
        coord_flip() +  # flip coordinates (puts labels on y axis)
        xlab("") + ylab("predicted % difference with cranial lesions present (95% CI)") +
        theme_bw() 

save_plot(bloodforest, base_width = 4.5, filename = "./Figures/posterior_deltas_blood_forestplot.pdf")


save_plot(raw_plots, base_width = 4.5, filename = "./Figures/Supplemental_Figures/S_raw_lesion_counts.pdf")
bloodforest
```


